---
title: "Data Transformation Columns"
echo: TRUE
format:
  revealjs: 
    theme: dark
    code-line-numbers: false
---

## **Learning objectives:**

-   **Pick out columns** of a data frame with **`dplyr::select()`**.

-   **Modify columns** of a data frame with **`dplyr::mutate()`**.

-   **Group rows** of a data frame with **`dplyr::group()`**.

-   **Apply functions to columns** of a (grouped) data frame with **`dplyr::summarize()`**.

-   **Conditional Formatting** of data

    ```{r}
    #| echo: false
    #| warning: false
    #| message: false

    library(tidyverse)
    library(nycflights13)

    ```

## Verbs

There are four important verbs that affect the columns without changing the rows:

-   `mutate()` creates new columns that are derived from the existing columns
-   `select()` changes which columns are present
-   `rename()` changes the names of the columns
-   `relocate()` changes the positions of the columns.

## Add new variables with mutate() {.unnumbered}

`mutate()` adds new columns based on values from existing columns. Data frame includes existing and new columns.

Convert the arrival delay and departure delay times, currently in minutes, into hours:

```{r}
#| code-line-numbers: "2,3,4,5|2,5|3|4|2,3,4,5"
flights |> 
  mutate(
    dep_delay_hrs = dep_delay/60,
    arr_delay_hrs = arr_delay/60
  )
```

## Select columns with select() {.unnumbered}

`select()` lets you pick which columns (variables) to use. The first argument to `select()` is a data frame, the subsequent arguments are columns to use.

The order you list the columns will determine the order of the columns returned by `select()`.

## Select columns by name {.unnumbered}

```{r}
# Select columns by name:
flights |> 
  select(year, month, day)
```

## Select columns by range {.unnumbered}

```{r}
# Select all columns between year and day (inclusive):
flights |> 
  select(year:day)
```

## Select columns via exclusion {.unnumbered}

-   This is equivalent to dropping columns you do not need

```{r}
# Select all columns except those from year to day (inclusive):
flights |> 
  select(!year:day)
```

## Select column by type {.unnumbered}

```{r}
# Select all columns that are characters:
flights |> 
  select(where(is.character))
```

## Helper functions for `select()`

There are a number of helper functions you can use within `select()`:

-   `starts_with("abc")`: matches names that begin with “abc”.
-   `ends_with("xyz")`: matches names that end with “xyz”.
-   `contains("ijk")`: matches names that contain “ijk”.
-   `num_range("x", 1:3)`: matches x1, x2 and x3.

See `?select` for more details.

## Select and rename columns

You can rename variables as you `select()` them by using `=`. The new name appears on the left hand side of the `=`, and the old variable appears on the right hand side:

```{r}
flights |> 
  select(tail_num = tailnum)
```

## Rename columns with rename() {.unnumbered}

If you want to keep all the existing variables and just want to rename a few, you can use `rename()` instead of `select()`:

```{r}
flights |> 
  rename(tail_num = tailnum)
```

**Tip**: If you have inconsistently named columns and it's difficult to fix them manually, use `janitor::clean_names()`. It automates the cleaning process.

## The pipe {.unnumbered}

The real power of the pipe is to combine multiple verbs.

Example: To find the fast flights to Houston’s IAH airport: you need to combine filter(), mutate(), select(), and arrange():

```{r }
#| code-line-numbers: "1,2,3,4,5|1|2|3|4|5|1,2,3,4,5"
flights |> 
  filter(dest == "IAH") |> 
  mutate(speed = distance / air_time * 60) |> 
  select(year:day, dep_time, carrier, flight, speed) |> 
  arrange(desc(speed))
```

## Piping for clarity {.unnumbered}

Using the pipe operator makes complex operations easier to read and follow. Each step is clear as the verbs start at the beginning of the line. Without the pipe, functions would be nested inside each other, making the code harder to understand. Think Excel nested stamentents!

```{r}
#| eval: false
#| code-line-numbers: "1,2,3,4,5,6,7,8,9,10,11,12,13|4,5,6,7|3,4,5,6,7,8,9|2,3,4,5,6,7,8,9,10,11|1,2,3,4,5,6,7,8,9,10,11,12,13"
arrange(
  select(
    mutate(
      filter(
        flights, 
        dest == "IAH"
      ),
      speed = distance / air_time * 60
    ),
    year:day, dep_time, carrier, flight, speed
  ),
  desc(speed)
)
```

## What if we didn't pipe

Using intermediate objects:

```{r}
#| eval: false
#| code-line-numbers: "1,2,3,4,5|1|2|3|4|5|1,2,3,4,5"
flights1 <- filter(flights, dest == "IAH")
flights2 <- mutate(flights1, speed = distance / air_time * 60)
flights3 <- select(flights2, year:day, dep_time, carrier, flight, speed)
arrange(flights3, desc(speed))
```

Both forms have their uses, but the pipe operator usually makes data analysis code easier to write and read.

**Tip**: To add the pipe to your code, use the built-in keyboard shortcut Ctrl/Cmd + Shift + M. (note: this may type the "old" pipe `%>%` which works just as well)

## Groups {.unnumbered}

The previous functions work with rows and columns. `dplyr` also allows to work with groups with functions like: `group_by()`, `summarize()`, and the slice family of functions.

### group_by() {.unnumbered}

Use `group_by()` to divide your dataset into groups meaningful for your analysis:

```{r}
flights |> 
  group_by(month)
```

group_by() doesn’t change the data but looking at the output, it indicates that it is “grouped by” month `(Groups: month [12])`

## summarize() {.unnumbered}

Summarization is the most important grouped operation, because it reduces the data frame to a single row for each group, summarizing the data with a single statistic. The following example computes the average departure delay by month:

```{r}
#| code-line-numbers: "1,2,3,4,5|1|2|3,4,5|1,2,3,4,5"
flights |> 
  group_by(month) |> 
  summarize(
    avg_delay = mean(dep_delay)
  )

```

The above example show NAs because some of the observed flights had missing data in the delay column. To ignore the missing data, use:

```{r}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE)
  )
```

In a single call to summarize() you can create any number of summaries. There are different useful summaries available, one of which is n(), which returns the number of rows in each group:

```{r}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE), 
    n = n()
  )
```

## Slice functions {.unnumbered}

To extract specific rows within each group you can use the following functions:

-   `df |> slice_head(n = 1)` takes the first row from each group.
-   `df |> slice_tail(n = 1)` takes the last row in each group.
-   `df |> slice_min(x, n = 1)` takes the row with the smallest value of column `x`.
-   `df |> slice_max(x, n = 1)` takes the row with the largest value of column `x`.

## Slice example

You can use n to select multiple rows or use prop = 0.1 to select 10% of the rows in each group. For example, the code finds the most delayed flights upon arrival at each destination.

```{r}
flights |> 
  group_by(dest) |> 
  slice_max(arr_delay, n = 1) |>
  relocate(dest)
```

## Slice Example

In the previous output we get 108 rows but there are only 105 destinations. `slice_min()` and `slice_max()` keep tied values so `n = 1` means give us all rows with the highest value, if you want only one row per group you can set `with_ties = FALSE`.

## Grouping by multiple variables {.unnumbered}

To make a group for each date you can create groups using more than one variable.

```{r}
daily <- flights |>  
  group_by(year, month, day)
daily
```

## Grouping by multiple variables {.unnumbered}

When you summarize a tibble grouped by more than one variable, each summary peels off the last group.

```{r}
daily_flights <- daily |> 
  summarize(n = n())
```

## Ungrouping {.unnumbered}

You might also want to remove grouping from a data frame without using `summarize()`. You can do this with `ungroup()`.

```{r}
daily |> 
  ungroup()
```

## Ungrouping Example

Now let's see what happens when you summarize an ungrouped data frame.

```{r}
daily |> 
  ungroup() |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE), 
    flights = n()
  )
```

You get a single row back because dplyr treats all the rows in an ungrouped data frame as belonging to one group.

## Conditional Formatting

Aim: generate a vector where each element is determined by the value of one or multiple conditions (\~ comparisons).

-   one condition: use `if_else()`
    -   `if_else(condition, true, false, missing = NULL)`
-   multiple (hierarchical) conditions: use `case_when(..., .default = NULL)`.
    -   the *first* condition that is `TRUE` for an element determines the outcome for that element.

## Conditional transformations (2) {.unnumbered}

```{r}
daily_otr <- daily |>
             mutate(dep_delay_hi = ifelse(dep_delay > 120, "Yes", "No"))
```

## Conditional transformations (3) {.unnumbered}

```{r}
daily_otr <- daily |>
             mutate(dep_delay_hi_mult = case_when(
               dep_delay < 0 ~ "Early",
               dep_delay == 0 ~ "On Time",
               dep_delay < 120 ~ "Low Delay",
               dep_delay >= 120 ~ "High Delay"
               ))
```

## Conditional transformations (4) {.unnumbered}

The different outcomes must be **compatible types**!

Some examples:

-   numerical and logical
-   strings and factors.
-   `NA` is compatible with everything.
