---
title: "Visualization Deep Dive  Exploratory Data Analysis"
echo: TRUE
format:
  revealjs: 
    theme: dark
    code-line-numbers: false
---

## Learning objectives

```{r}
#| echo: false
library(tidyverse)
```

-   Recognize the **two types of questions** that will always be useful for making discoveries within your data: "What type of variation occurs within my variables?" and "What type of covariation occurs between my variables?"
-   Explore the **variation** within the variables of your observations.
-   Deal with outliers and **missing values** in your data.
-   Explore the **covariation** between the variables of your observations.
-   Recognize how **models** can be used to explore **patterns** in your data.

## Overall Vocabulary

-   **variable:** a quantity, quality, or property that you can measure.
-   **value:** the state of a variable when you measure it. Can change.
-   **observation:** a set of measurements made under similar conditions. One value per variable.
-   **tabular data:** observations of variables.
-   **tidy data:** 1 observation per row, 1 variable per column, 1 value per cell. Definition of "tidy" for a dataset can depend on what you're trying to answer.

## Remember

-   the `|>` pipe = "and then".
-   `{ggplot2}` uses `+` to add layers, read it as "with" or "and".

## Variation

-   **variation:** the tendency of values of a variable to change between measurements.

## Variation in categorical variables

-   **categorical variable:** can only take certain values. Visualize variation with bar chart.

```{r}
ggplot(data = diamonds) +
  aes(x = cut) +
  geom_bar()
```

## Variation in continuous variables

-   **continuous variables:** can take on infinite set of ordered values. Visualize variation with histogram.

```{r}
ggplot(data = diamonds) +
  aes(x = carat) +
  geom_histogram(binwidth = 0.5)
```

## Why Visualize?

-   Use the visualizations to develop questions!
    -   Which values are the most common? Why?
    -   Which values are rare? Why? Does that match your expectations?
    -   Can you see any unusual patterns? What might explain them?

## Why Group Data?

-   Subgroups create more questions:
    -   How are the observations within each cluster similar to each other?
    -   How are the observations in separate clusters different from each other?
    -   How can you explain or describe the clusters?
    -   Why might the appearance of clusters be misleading?

## Some notes:

-   Can be ok to drop weird values, especially if you can explain where they came from.
-   Always disclose that you did that, though.

## Missing values

2 options to deal with weird values:

-   Drop the entire row. \<-- probably don't do this
-   Replace odd data with NA.
-   `{ggplot2}` will give a warning when values are missing, can suppress with `na.rm = TRUE`.

## Covariation

-   **covariation:** tendency of values of different variables to vary *together* in a related way.
-   Visualizing covariance depends on types of variables in the pair

## Covariation categorical + continuous

-   `x = categorical`
-   `y = continuous`.
-   `geom_boxplot`
-   Lots of options exist to do this better. See [Cedric Scherer's tutorial](https://www.cedricscherer.com/2021/06/06/visualizing-distributions-with-raincloud-plots-with-ggplot2/)!

## Covariation categorical + categorical

-   `geom_count`
-   `dplyr::count` then `geom_tile`

## Covariation continuous + continuous

-   `geom_point`
-   `geom_bin2d`
-   `geom_hex`

## Remember

**Ask yourself:**

-   Could this pattern be due to coincidence (i.e. random chance)?
-   How can you describe the relationship implied by the pattern?
-   How strong is the relationship implied by the pattern?
-   What other variables might affect the relationship?
-   Does the relationship change if you look at individual subgroups of the data?

## Learning More

-   [dslc.io/join](dslc.io/join) for more book clubs!
-   [R Graph Gallery](https://www.r-graph-gallery.com/ggplot2-package.html)
-   The [Graphs section](http://www.cookbook-r.com/Graphs/) of the R Cookbook
