---
title: "Conditional Transformations"
echo: TRUE
format:
  revealjs: 
    theme: dark
    code-line-numbers: false
---

## Objectives {.unnumbered}

Aim: generate a vector where each element is determined by the value of one or multiple conditions (\~ comparisons).

-   one condition: use `if_else()`
    -   `if_else(condition, true, false, missing = NULL)`
-   multiple (hierarchical) conditions: use `case_when(..., .default = NULL)`.
    -   the *first* condition that is `TRUE` for an element determines the outcome for that element.

## Conditional transformations (2) {.unnumbered}

```{r}
x <- c(-3:3, NA)
if_else(x > 0, "+ve", "-ve", "???")
```

## Conditional transformations (3) {.unnumbered}

```{r}
x <- c(-3:3, NA)
case_when(
  x == 0   ~ "0",
  x < 0    ~ "-ve", 
  x > 0    ~ "+ve",
  is.na(x) ~ "???"
)
```

## Conditional transformations (4) {.unnumbered}

The different outcomes must be **compatible types**!

Some examples:

-   numerical and logical
-   strings and factors.
-   `NA` is compatible with everything.
